---
title: "class07"
author: "Maasfa Shaikh"
format: pdf
editor: visual
---

```{r}
# Generate some example data for clustering
tmp <- c(rnorm(30,-3), rnorm(30,3))
x <- cbind(x=tmp, y=rev(tmp))
plot(x)
```
```{r}
km<- kmeans(x, centers =2, nstart=20)
km

km$size
km$cluster
km$center
```
> There are 30 points in each cluster
> cluster size is answered by km$size
>cluster assignment is answered by km$cluster
>cluster center is answered by km$centers

*plotting x colored by kmeans cluster assignnment and adding cluster centers as blue points*

```{r}
plot(x, col = km$cluster)
#plotting clusters

points(km$centers, col = "blue", pch = 15, cex = 1)
#assigning the color blue to cluster centers
```

*using hclust to analyze data*

```{r}
d <- dist(x)
#computing distance matrix

hc <- hclust(d)
#doing heirarchical clusters using hclust and assigning it to hc

plot(hc)
#plotting dendrogram

groups <- cutree(hc, h=8)
#cutting the tree into 2 clusters using cutree, with k set to 2 

table(groups)
#checking cluster size

plot(x, col = groups, pch = 16)
#plotting with colored clusters
```

```{r}
plot(hc)
abline(h = 5, col = "red")

#cluster membership = groups

#cluster size = table (groups)
```











```{r}
url <- "https://tinyurl.com/UK-foods"
x <- read.csv(url)
```

Q1: How many rows and columns are in your new data frame named x? What R functions could you use to answer this questions?

```{r}
dim(x)

#there are 17 rows and 5 columns in the new data frame. I used the dim function to answer this.
```

```{r}
#checking data

head(x)
```

```{r}
#making it so that the names dont count as a column

rownames(x) <- x[,1]
x <- x[,-1]
head(x)

dim(x)
```

```{r}
#second method of solving row name problem

x <- read.csv(url, row.names=1)
head(x)

dim(x)
```

Q2. Which approach to solving the ‘row-names problem’ mentioned above do you prefer and why? Is one approach more robust than another under certain circumstances?

> A2: I like the second approach better because it is faster to do the code in one line from the start instead of writing multiple codes. I also found that running the first code repeatedly got rid of more columns

*Spotting major differences and trends*

```{r}
barplot(as.matrix(x), beside=T, col=rainbow(nrow(x)))

```

Q3: Changing what optional argument in the above barplot() function results in the following plot?

> A3: changing beside to false changed the plot.It made it so that the bars of the same country are not next to eachother.

```{r}
barplot(as.matrix(x), beside=F, col=rainbow(nrow(x)))
```

```{r}
#converting to long format for ggplot with pivot longer()

library(tidyr)

x_long <- x |> 
          tibble::rownames_to_column("Food") |> 
          pivot_longer(cols = -Food, 
                       names_to = "Country", 
                       values_to = "Consumption")

dim(x_long)

```

```{r}
#making ggplot
library(ggplot2)

ggplot(x_long) +
  aes(x = Country, y = Consumption, fill = Food) +
  geom_col(position = "dodge") +
  theme_bw()
```

Q4: Changing what optional argument in the above ggplot() code results in a stacked barplot figure?

> A\$: changing the "dodge" argument to "stack" in geom_col gave me a plot that was the same as the one in the worksheet

```{r}
ggplot(x_long) +
  aes(x = Country, y = Consumption, fill = Food) +
  geom_col(position = "stack") +
  theme_bw()
```

Q5: We can use the pairs() function to generate all pairwise plots for our countries. Can you make sense of the following code and resulting figure? What does it mean if a given point lies on the diagonal for a given plot? code: pairs(x, col=rainbow(nrow(x)), pch=16)

> A5: This code basically plots every country against eachother. X represents the data set, col=rainbow(nrow(x)) gives each country a different color. pch=16 means that there are 16 solid points in each graph. Each point represents a country and each panel represents the relationship between two variables. If a point lies on a diagnol of a plot, it means its not special or unusual, the country's value is identical on both axes.

*heat map*

```{r}
library(pheatmap)

pheatmap( as.matrix(x) )
```

Q6. Based on the pairs and heatmap figures, which countries cluster together and what does this suggest about their food consumption patterns? Can you easily tell what the main differences between N. Ireland and the other countries of the UK in terms of this data-set?

> wales and England cluster together very tightly, and then scotland. This shows high correlation in diet amongst UK countries. Ireland is a little more off to the side compared to the rest. Ireland differs in a combination of many food categories, for example, Ireland has the highest alcohol consumption compared to the other countries. 


*performing PCA*

```{r}
pca <- prcomp( t(x) )
summary(pca)

```
Q7. Complete the code below to generate a plot of PC1 vs PC2. The second line adds text labels over the data points.

```{r}
# Create a data frame for plotting
df <- as.data.frame(pca$x)
df$Country <- rownames(df)

# Plot PC1 vs PC2 with ggplot
ggplot(pca$x) +
  aes(x = PC1, y = PC2, label = rownames(pca$x)) +
  #added PC1 for x and PC2 for y
  geom_point(size = 3) +
  geom_text(vjust = -0.5) +
  xlim(-270, 500) +
  xlab("PC1") +
  ylab("PC2") +
  theme_bw()
```


Q8. Customize your plot so that the colors of the country names match the colors in our UK and Ireland map and table at start of this document.

```{r}
df <- as.data.frame(pca$x)
df$Country <- rownames(df)
country_colors <- c(
  "England" = "orange",
  "Scotland" = "blue",
  "Wales" = "red",
  "N.Ireland" = "green"
)
#made a table with the colors I want the countries to be



# Plot PC1 vs PC2 with ggplot
ggplot(pca$x) +
  aes(x = PC1, y = PC2, label = rownames(pca$x), color= df$Country) +
  #colored the countries for the plot
  geom_point(size = 3) +
  geom_text(vjust = -0.5) +
  scale_color_manual(values = country_colors)+
  #added the color scale I wanted by assigning the color vector to the scale
  xlim(-270, 500) +
  xlab("PC1") +
  ylab("PC2") +
  theme_bw()
```


```{r}
v <- round( pca$sdev^2/sum(pca$sdev^2) * 100 )
v
z <- summary(pca)
z$importance
```
```{r}
# Create scree plot with ggplot
variance_df <- data.frame(
  PC = factor(paste0("PC", 1:length(v)), levels = paste0("PC", 1:length(v))),
  Variance = v
)

ggplot(variance_df) +
  aes(x = PC, y = Variance) +
  geom_col(fill = "steelblue") +
  xlab("Principal Component") +
  ylab("Percent Variation") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 0))
```

```{r}
## Lets focus on PC1 as it accounts for > 90% of variance 
ggplot(pca$rotation) +
  aes(x = PC1, 
      y = reorder(rownames(pca$rotation), PC1)) +
  geom_col(fill = "steelblue") +
  xlab("PC1 Loading Score") +
  ylab("") +
  theme_bw() +
  theme(axis.text.y = element_text(size = 9))

```
Q9: Generate a similar ‘loadings plot’ for PC2. What two food groups feature prominantely and what does PC2 maninly tell us about?

> This loading plot using PC2 features soft drinks as the largest positive loading score, and fresh potatoes as the highest negative score. PC2 tells us which foods define the second major direction of variation in the data.  

```{r}
ggplot(pca$rotation) +
  aes(x = PC2, 
      y = reorder(rownames(pca$rotation), PC2)) +
  geom_col(fill = "steelblue") +
  xlab("PC2 Loading Score") +
  ylab("") +
  theme_bw() +
  theme(axis.text.y = element_text(size = 9))
```


> **Section 2 optional extra credit**

```{r}
url2 <- "https://tinyurl.com/expression-CSV"
rna.data <- read.csv(url2, row.names=1)
head(rna.data)

dim(rna.data)
nrow(rna.data)
```

Q10: How many genes and samples are in this data set? How many PCs do you think it will take to have a useful overview of this data set (see below)?

> A10: There are 100 genes in this dataset and 10 samples. We will probably need only 1 pcs to have a useful overview of this data set because 1 pc retained 92.62% variance, meaning it is that similar to the original dataset. 

```{r}
## Again we have to take the transpose of our data 
pca <- prcomp(t(rna.data), scale=TRUE)

# Create data frame for plotting
df <- as.data.frame(pca$x)
df$Sample <- rownames(df)

## Plot with ggplot
ggplot(df) +
  aes(x = PC1, y = PC2, label = Sample) +
  geom_point(size = 3) +
  geom_text(vjust = -0.5, size = 3) +
  xlab("PC1") +
  ylab("PC2") +
  theme_bw()

summary(pca)
```


```{r}
# Calculate variance explained
pca.var <- pca$sdev^2
pca.var.per <- round(pca.var/sum(pca.var)*100, 1)

# Create scree plot data
scree_df <- data.frame(
  PC = factor(paste0("PC", 1:10), levels = paste0("PC", 1:10)),
  Variance = pca.var[1:10]
)

ggplot(scree_df) +
  aes(x = PC, y = Variance) +
  geom_col(fill = "steelblue") +
  ggtitle("Quick scree plot") +
  xlab("Principal Component") +
  ylab("Variance") +
  theme_bw()
```


```{r}
## Percent variance is often more informative to look at 
pca.var.per

# Create percent variance scree plot
scree_pct_df <- data.frame(
  PC = factor(paste0("PC", 1:10), levels = paste0("PC", 1:10)),
  PercentVariation = pca.var.per[1:10]
)

ggplot(scree_pct_df) +
  aes(x = PC, y = PercentVariation) +
  geom_col(fill = "steelblue") +
  ggtitle("Scree Plot") +
  xlab("Principal Component") +
  ylab("Percent Variation") +
  theme_bw()
```

```{r}
## A vector of colors for wt and ko samples
colvec <- colnames(rna.data)
colvec[grep("wt", colvec)] <- "red"
colvec[grep("ko", colvec)] <- "blue"

# Add condition to data frame
df$condition <- substr(df$Sample, 1, 2)
df$color <- colvec

ggplot(df) +
  aes(x = PC1, y = PC2, color = color, label = Sample) +
  geom_point(size = 3) +
  geom_text(vjust = -0.5, hjust = 0.5, show.legend = FALSE) +
  scale_color_identity() +
  xlab(paste0("PC1 (", pca.var.per[1], "%)")) +
  ylab(paste0("PC2 (", pca.var.per[2], "%)")) +
  theme_bw()
```

```{r}
loading_scores <- pca$rotation[,1]

## Find the top 10 measurements (genes) that contribute
## most to PC1 in either direction (+ or -)
gene_scores <- abs(loading_scores) 
gene_score_ranked <- sort(gene_scores, decreasing=TRUE)

## show the names of the top 10 genes
top_10_genes <- names(gene_score_ranked[1:10])
top_10_genes 
```










